Human capital mediates natural selection in contemporary humans

### 2021-12-15 why-natural-selection.Rmd
More of a mediation analysis

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-educ)), with
large reductions for polygenic scores for educational attainment. This 
suggests that education indeed mediates the effect of polygenic scores on fertility.

@@

\FloatBarrier
\clearpage
## Controlling for earnings and education

```{r plot-res-educ, fig.cap = "Selection effects controlling for education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison."}

drake::loadd(res_ee_control)
drake::loadd(res_all)

res_ee_control %<>% filter(term == score_name)

res_all %<>% 
  filter(term == score_name, reg.type == "controlled", dep.var == "N children") %>% 
  select(-dep.var, -reg.type)

res_compare <- bind_rows(
                 Raw = res_all, 
                 Controlled = res_ee_control, 
                 .id = "Regression"
               )

standard_ggplot(res_compare, fill_col = Regression, n_regs = 33)

@@

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}}=\frac{\sigma_{p}^{2}-\sigma_{\eta}^{2}}{\sigma_{p}^{2}}=\frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}\equiv\lambda,\label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.

>>>>

Second we test whether educational attainment mediates the correlation between
polygenic scores and fertility.

@@

\FloatBarrier
\clearpage
## Mediation analysis

We run a standard mediation analysis in the framework of @baron1986moderator. We estimate 
\begin{align}
N_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
\end{align}
where $N_i$ is fertility, $PGS_i$ is each polygenic score, $EA_i$ is educational
attainment (age of leaving fulltime education), and $X_i$ is a vector of
controls. The total effect of $PGS$ on number of children is $\beta + \gamma
\zeta$. The "indirect effect" mediated by $EA$ is $\gamma \zeta$. The standard
error of the indirect effect can be calculated as

\[
\sqrt{\hat\gamma^2 \hat\sigma_\zeta^2 + \hat\zeta^2\hat\sigma_\gamma^2}
\]

where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc.

We are not concerned about endogeneity of polygenic scores, since our theory
requires only that these correlate with human capital. However, endogeneity of
educational attainment is a concern. For example, educational attainment might
correlate with unobserved characteristics that affect one's fertility, e.g. via
success in marriage markets. To partially address this, we include
controls for age, sex, fluid IQ and height in $X$.

```{r tbl-res-mediation}

drake::loadd(res_mediation)

res_mediation %<>% mutate(
                     prop = estimate_ind/estimate_total,
                     # 0.025 not 0.05 because two-sided
                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
                     star = ifelse(sig, "*", "")
                   )

n_positive <- sum(res_mediation$prop > 0)
n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
median_prop <- median(res_mediation$prop)

res_mediation %>%
      arrange(desc(estimate_total)) %>%
      transmute(
        PGS = pretty_names(term),
        `Total effect` = estimate_total,
        `Indirect effect` = estimate_ind,
        `95% c.i.` = glue::glue(
                     "[{estimate_ind - 1.96 * se_ind}, {estimate_ind + 1.96 * se_ind}]"
                     ),
        `Indirect/total` = prop,
        ` ` = star
      ) %>%
      as_hux() %>%
      set_align(everywhere, -1, "right") %>% 
      set_number_format(-1, 2:4, 4) %>% 
      set_number_format(-1, 5, fmt_percent()) %>%
      set_caption("Mediation analysis") %>% 
      theme_compact() %>% 
      set_width(0.9) %>%
      set_col_width(c(.3, .125, .125, .3, .1, .05)) %>% 
      set_bottom_border(final(), everywhere) %>%
      add_footnote("* indirect effect p < 0.05/33", font_size = 8)
```

Table \@ref(tab:tbl-res-mediation) shows results. `r n_pos_sig` scores are 
significantly greater than 0 at $p = 0.05/33$.

@@

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
= \frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}
\equiv \lambda, \label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.
